Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2024-03-13 2_SSL_NE/article/article_main.tex
corrections for reviewer 1
updated figures and captions

\begin{figure}[h!]
\centering
  \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 7 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

\subsection{Evolution of LED Chip Architectures}

@@

\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies. The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}. Through the index decomposition analysis described in \cref{subsec:performance_metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to innovations that involved technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.

@@

\begin{figure}[H]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020} (top to bottom panels). Losses are normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.  Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10,  adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

@@

Historical improvements in consumer experience metrics for phosphor-converted warm white GaN-based LEDs are shown in \cref{fgr:consumer_experience}. In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable. Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into the white light required for general illumination . It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs. The first commercial white LED produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}. As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.

@@

Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.

@@

    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|X|l|X|}

@@

        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
        1926 & 1994 & 1996 & LED phosphors & Use of phosphors for light down conversion in LEDs & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline

@@

According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
%
\begin{align}
    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
\end{align}
%
where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$. These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies: 
%
\begin{equation}
    \Delta a + \Delta b  = \Delta F
\end{equation}
%
To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered. However, the scope of data reporting in the literature is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance. Where gaps in relevant data existed, we filled them with our own performance calculations for individual sub-efficiencies where possible. Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python \cite{colour-science_software}. We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 4 and Supplementary Figure 10 for the associated spectral data and calculation results). This approach allowed us to quantify the improvements related to consumer experience and phosphor development in LEDs.

@@

The structure of our bottom-up white LED manufacturing cost model with process-step resolution is generally based on the 2012 \textit{LEDCOM} cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends. The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts. For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive , and expert interviews. Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Figure 3 and Supplementary Figures 2-6 for the details of each architecture). The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks  and relevant patents. In addition, two LED life cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs. These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.

>>>>

\begin{figure}[h!]
\centering
  \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 7 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

% should be integrated into some other sub-heading
% \subsection{Evolution of LED Chip Architectures}

@@

@@

\begin{figure}[H]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020}. Losses are normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.  Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10, adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

@@

Historical improvements in consumer experience metrics for phosphor-converted warm white GaN-based LEDs are shown in \cref{fgr:consumer_experience}. In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable. Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into the white light required for general illumination . It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs. The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that could generate only cool white light \cite{bando1998development}. As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.

@@

Our findings are further supported by a preliminary sensitivity analysis, presented in Section 5 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.

@@

    \begin{NiceTabularX}{1.1\textwidth}{|l|l|l|X|X|X|X|X|X|}

@@

        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Enabler} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
        1926 & 1994 & 1996 & LED phosphors &  Use of phosphors for light down conversion in LEDs & ??? & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & ??? & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & ??? &  Chemistry (S), Materials science (S) & \cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T)  & \cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & research conference & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & ??? & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & ??? & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & ??? & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline

@@

According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
%
\begin{align}
    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
\end{align}
%
where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$. In practice, this mean can be approximated, simplifying calculations. These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies: 
%
\begin{equation}
    \Delta a + \Delta b  = \Delta F
\end{equation}
%
To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered. However, the scope of data reporting in the literature is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance. Where gaps in relevant data existed, we filled them with our own performance calculations for individual sub-efficiencies where possible. Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python \cite{colour-science_software}. We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 4 and Supplementary Figure 10 for the associated spectral data and calculation results). This approach allowed us to quantify the improvements related to consumer experience and phosphor development in LEDs.

@@

The structure of our bottom-up white LED manufacturing cost model with process-step resolution is generally based on the 2012 \textit{LEDCOM} cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends. The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts. For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive, and expert interviews. Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Figure 3 and Supplementary Figures 2-6 for the details of each architecture). The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents. In addition, two LED life-cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs. These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
